Flooding of the Okavango Delta influences Connectivity for Dispersing African Wild Dogs

### 2023-01-08 Chapter_8/04_Manuscript/Manuscript.tex
upd

\section{Introduction}
Climate change is expected to profoundly impact ecosystems worldwide with
far-reaching consequences for species living therein \citep{Ozgul.2010,
Radchuk.2019, IPCC.2022}. By altering environmental conditions, climate change
affects animal behavior \citep{Fuller.2016}, resource availability
\citep{Durant.2007}, population dynamics \citep{Paniw.2021}, and the
distribution of wild living animal populations \citep{Thomas.2004,
Thuiller.2006}. One life-history pathway through which species potentially
mediate the effects of environmental change is dispersal \citep{Anderson.2012},
i.e. the movement of individuals away from their natal location to the site of
first reproduction \citep{Clobert.2012}. The ability to disperse enables species
to adapt to climate change by tracking favorable habitat conditions
\citep{Raia.2012} or by shifting into a different region of their fundamental
niches \citep{Kokko.2022}. Dispersal furthermore provides resilience against
changing conditions by safeguarding genetic diversity \citep{Frankham.2002,
Leigh.2012, Baguette.2013}, facilitating the colonization empty habitats
\citep{Gustafson.1996, Hanski.1999b, MacArthur.2001}, and promoting the
reinforcement of weakened and small subpopulations \citep{Brown.1977}.

While dispersal offers a means to offset the demographic consequences of climate
change \citep{Kokko.2006, Hodgson.2009, Travis.2013}, it itself is a function of
climatic and environmental conditions (e.g. \citep{Elliot.2014, Behr.2020}). The
link between dispersal and the environment can either be indirect, for example
if the propensity of individuals to disperse depends on environmental
conditions, or direct, when the biophysical environment through which dispersers
move affects dispersal prospects \citep{Travis.2013}. The latter case highlights
that dispersal is also inextricably linked to the concept landscape connectivity
\citep{Baguette.2013}, which is understood as the degree by which the landscape
facilitates or impedes movements \citep{Taylor.1993}. A sufficient degree of
landscape connectivity is therefore a critical prerequisited of successful
dispersal \citep{Fahrig.2003}. Unfortunately, the continued degradation and
destruction habitats worldwide continues to imperil the dispersal ability of
many species and increases the prospects of inbreeding, genetic risk and of
local extirpation due to stochastic events \citep{Melbourne.2008, Sawyer.2011}.
Combined, these factors limit species ability to adapt to changing environmental
conditions \citep{Fahrig.2003}. Conservation strategies that aim to facilitate
dispersal by improving landscape connectivity are thus often viewed as pinacle
of conservation strategies to combat the adverse effects of climate change
\citep{Heller.2009}.

In recent years, there has been a growing body of research that uses animal
relocation data to identify movement corridors and assess connectivity at large
scales \citep{Chetkiewicz.2006, Doerr.2011, Squires.2013, Elliot.2014,
Benz.2016, Osipova.2019}. Identification of potential corridors relies on the
estimation of permeability (respectively resistance) surfaces, which return the
ease or willingness at which the focal species traverses a specific landscape
\citep{Sawyer.2011}. Such surfaces are typically created on the basis of habitat
preferences, which can be estimated using a suite of selection functions
\citep{Boyce.2002, Fortin.2005, Cushman.2010}. In selection functions, habitat
preferences are inferred by comparing spatial covariates (e.g. environmental and
anthropogenic) at locations visited by the animal to the same spatial covariates
at randomly selected locations in the animal's availability domain
\citep{Zeller.2012, Thurfjell.2014}. Importantly, selection functions rely on
adequate landscape and GPS relocation data that are representative of the
process being studied \citep{Diniz.2020}. More specifically, GPS relocation data
collected on dispersing individuals has been shown to outperform GPS relocation
data collected on resident individuals when aiming at detecting large scale
dispersal corridors \citep{Elliot.2014, Diniz.2020}. However, few connectivity
studies have partitioned their GPS relocation data according to behavioral modes
(e.g. resident vs. dispersing) of the studied species \citep{Wilson.2012,
Vasudev.2015}. As such, most permeability surfaces upon which dispersal
corridors are identified are created using GPS relocation data that were
collected on resident individuals. This introduces severe biases and
substantially reduces the power to reveal meaningful movement corridors, for
dispersing individuals have different needs and drives compared to resident
individuals \citep{Killeen.2014, Elliot.2014, Cozzi.2020}. Consequently, these
biases have limited our ability to assess the adequacy of the boundaries of
protected areas in securing connectivity for their protected species.

To study connectivity, various modeling techniques have emerged
\cite{Etherington.2016, Diniz.2019}. Initially, the techniques were limited to
studying strucutral aspects of connectivity by focusing on the composition and
configuration of habitat patches, while ignoring species' responses to the
landscape matrix \citep{Tischendorf.2000, Doerr.2011}. With the increasing
availability of telemetry data and methods to study habitat and movement
preferences \citep{Boyce.2002, Fortin.2009, Cushmann.2010, Avgar.2016}, however,
the focus has shifted towards more functional aspects of connectivity
\citep{Tischendorf.2000, Doerr.2011}. Currently, the most prominent
\textit{functional} connectivity models are least-cost path analysis (LCPA,
\citealp{Adriaensen.2003}) and circuit theory (CT, \citealp{McRae.2008}), two
graph-based methods that estimate conductance of the landscape by means of a
resistance (or inversely permeability) surface \citep{Zeller.2012}. Because both
approaches make assumptions that are rarely met by dispersing individuals,
individual-based movement models (IBMMs) have also gained some momentum recently
\citep{Kanagaraj.2013, Allen.2016, Hauenstein.2019, Diniz.2019, Zeller.2020,
Unnithan.2021, Unnithan.2022, Hofmann.2022}. In these models, movement of
individuals is explicitly simulated and connectivity metrics are derived based
on the simulated dispersal paths. IBMMs provide great modeling flexibility and
are thus considered powerful tools for examining connectivity under different
landscape configurations \citep{Littlefield.2019}.

Irrespective of the chosen method, connectivity is often modeled as a snapshot
in time that fails to render dynamic environmental change. However, some recent
studies have shown that connectivity may differ substantially depending on
season ... some examples.

 Unfortunately, our
understanding how changing environmental conditions influence connectivity and
dispersal is limited \citep{Kokko.2006, Travis.2013}. Because collecting
information on dispersing animals is difficult and because it is difficult to
assess how changing environmental conditions affect connectivity...

\cite{Ashrafzadeh.2019} found that cimate change is likely to substantially
reduce connectivity for mountain newts in Iran. Similarly, \cite{Luo.2021}
predicted significant reductions in connectivity for the giant spiny frog in
South-East Asia.

However, failing to take into account potential changes in climatic and
environmental conditions imperils our ability to design cost-effective
conservation strategies.

Here, we used naturally occuring seasonal flood-cycles to approximate the likely
changes in connectivity that will arise under amplified climatic conditions due
to climate change. While a better understanding of the ODs flooding regime under
global change is of major interest, reliable predictions remain challenging due
to the complex interplay between climate, anthropogenic water use, animal
behavior, and topography of the region.

Climate change
could lead to the disruption of existing dispersal corridors and the emergence
of new dispersal barriers but may also provide possibilities to access
previously unsuitable dispersal routes. Designing and implementing effective
conservation strategies that aim to improve landscape connectivity, such as
habitat restoration or the creation of wildlife corridors, requires better
understanding of such dynamics.

Albeit the impacts of climate change on connectivity cannot be directly
estimated, one can study how changing environmental conditions due to
seasonality influence patterns of dispersal and connectivity.

Habitat degradation and fragmentation lead to isolated subpopulations that have
a high associated risk of extinction due to stochastic events and generally
decreased genetic variability \citep{Melbourne.2008}.

More recently, individual-based movement models (IBMMs) have been uprising and
are recommended for their ability to incorporate several crucial realities of
animal behavior, such as, for example, autocorrelation, that are by omitted by
necessity using resistance-based approaches \citep{Hofmann.2023,
Unnithan.2022a}.
Botswana is among the countries most heavily influenced by climate change .

@@

Earlier studies demonstrating the importance of rendering seasonal change
include \cite{Chetkiewicz.2009}, who use resource selection functions to demonstrate
that connectivity differs for Ursus actor and Puma concolor and \cite{Mui.2017},
who showed that connectivity for turtles changes depending on the season, and
\cite{Kaszta.2021}, who show monthly differences in elephant's habitat selection
using path-selection functions.

Only if there

@@

The Okavango delta in Southern Africa poses a unique opportunity to study the
impacts of environmental change on species dispersal ability and connectivity in
a large scale natural experiment setup. According to projections by the IPCC,
Botswana and its surroundings are among the most vulnerable to climate change.
While global temperatures are expected to increase by xx degrees until the end
of the 21st century, For example, \cite{Engelbrecht.2015} predicts an increase
between 4 and 6\degree C for regions in southern Africa. One of the keyspone
predators and an umberella species for conservation efforts in this ecosystem is
the African wild dog (\textit{Lycaon pictus}). This species heavily relies on
dispersal for long-term persistence. The African wild dog is a lightly built
canid lives in cohesive packs comprising up to 20 individuals. Wild dogs are
social breeders and the majority of reproduction is monopolized by the dominant
alpha couple who are supported by subordinates in taking care of their pups. The
species exhibits strong inbreeding avoidance, hence juvenile individuals born
into the pack disperse in same-sex sibling coalitions after reaching sexual
maturity \citep{McNutt.1996, Behr.2020}. Notably, the African wild dog is the
sole extant representative of its genus and thus considered as high-priority
species for conservation efforts \citep{Leigh.2012}. However, fragmentation and
destruction of their historic habitat may have decreased dispersal success for
this species and thus jeopardizes connectivity \citep{Leigh.2012}. Since the
impacts of climate change are not uniformly distributed across the planet, but
particularly pronounced at high latitudes and arid ecosystems, the species may
be seen as most vulnerable to such ecosystem shifts.

Talk about Zeller und Osipova

Here, we capitalize on the previously parameterized dispersal model by
\cite{Hofmann.2022} and use dispersal simulations to investigate dispersal
success and connectivity patterns for African wild dogs under two extreme
scenarios; one assuming maximum flooding of the Okavango delta and one assuming
minimum flooding of the Okavango delta. Given that water poses a barrier to
dispersing individuals, we anticipated that dispersal prospects and connectivity
during maximum flood are low. During minimum flood, in contrast, we expect to
reveal the presence of several dispersal corridors that may be used to freely
and quickly move between remaining habitat patches.

@@

\subsection{Spatial Habitat Layers}
We represented the physical landscape through which dispersers could move by a
set of spatially referenced habitat layers, each resolved at 250m x 250m. The
set of layers included water-cover, distance-to-water, tree-cover,
shrub/grassland-cover, and a human influence layer depicting anthropogenic
influences through villages, roads, and agriculture. A detailed description of
the different habitat layers is provided in \Cref{Hofmann.2021}. Importantly,
the water-cover and derived distance-to water layers were generated using MODIS
Terra MCD43A4 satellite imagery that was classified using a ``floodmapping''
algorithm developed by \citep{Wolski.2017} and available through the R-package
\textsf{floodmapr}. The algorithm allowed us to generate almost weekly updated
``floodmaps'', thus providing detailed information about the flood-extent at any
given point in time. In total, we generated 700 floodmaps between the years 2000
and 2019. Based on these maps, we generated a minimal and maximum flood
scenario. To create the minimum flood scenario, we averaged the 50 floodmaps
with smallest flood extent and generated a binary image using areas that were
inundated in at least 50\% of the maps. Similarly, we created an average image
for high flood using the 50 most flooded maps. The final maps are depicted in
\Cref{FloodExtent}.

@@

\subsection{Dispersal Model}
Our dispersal model was based on a previously parametrized and validated
integrated step-selection function (iSSF, \citealp{Avgar.2016}) fitted to GPS
data of dispersing AWDs \citep{Hofmann.2021}. In step selection functions (SSFs,
\citealp{Fortin.2005}), observed GPS locations are converted into steps (the
straight-line traveled between two GPS recordings \citep{Turchin.1998a}) and
compared to a set of \textit{random} steps in a (mixed effects) conditional
logistic regression framework \citep{Fortin.2005, Thurfjell.2014, Muff.2020,
Fieberg.2021}. The model presented in \citep{Hofmann.2021} used dispersal data
of 16 dispersing African wild dogs from a free-ranging wild dog population in
northern Botswana. GPS data during dispersal was collected at 4-hourly intervals
and translated into steps of similar duration. Observed steps were then paired
step with 24 random steps that were generated using a uniform distribution for
turning angles (\(-\pi, +\pi\)) and step lengths from a gamma distribution
fitted to observed steps (scale \(\theta\) = 6'308 and shape \(k\) = 0.37). It
was then assumed that animals assigned to each observed and random step a
selection score of the form \citep{Fortin.2005}:

@@

\subsection{Source Areas and Emigration Zones}
We simulated dispersing AWDs originating from six distinct source areas located
in the vicinity of the Okavango delta (\Cref{StudyArea}). For areas one to six
we selected locations at the delta's periphery that remained dry in both
scenarios. At those locations, we generated circular buffers with a radius of 20
km. For source area six, we isolated a polygon covering Chief's Island, a
peninsula located at the OD's center. Besides source areas, we also generated
``zones of emigration'' that we used as checkpoints to determine if and where
simulated individuals left the delta's vicinity (\Cref{StudyArea}). We generated
emigration zones by first overlaying the OD with an elliptic buffer zone that we
dissected using a set of cutlines that originated from the ODs center and spread
according to cardinal points (\Cref{StudyArea}).

\subsection{Dispersal Simulation}
For each source area we simulated 1'000 individuals, once assuming a minimum
flood, once assuming a maximum flood. This resulted in the simulation of 6'000
individuals for each environmental scenario, hence 12'000 individuals in total.
The simulation algorithm was based on the algorithm described in
\Cref{Hofmann.2022} and works as follows. A random location within the source
area is chosen as a starting point. Originating from the starting point, a set
of 25 random steps is generated by sampling step lengths from a gamma
distribution fitted to observed steps (what is a step?) (shape = , scale = ) and
turning angles from a uniform distibution ($-\pi, +\pi$). Along each random step
the underlying spatial covariates are extracted and relevant movement metrics
are computed (i.e. log(sl), cos(ta), ta). The parametrized dispersal model is
used to predict the probability of each step for being chosen given the steps
covariate values. One of the steps is sampled based on assigned probabilities
and the location of the animal is updated. The procedure is then repated until
the desired number of steps is realized. Here, we simulated each individual for
2'000 steps, which correspons to the longest dispersal duration recorded in our
data. The original model was trained using 4-hourly steps, thus a simulated step
also resembled the movement conducted within four hours. Trajectories resulting
from such a simulation can be understood as correlated random walks that take
into account both habitat and movement preferences of dispersing individuals.

@@

\subsection{Dispersal Prospects and Connectivity}
Based on simulated dispersal trajectories in the two scenarios we quantified
dispersal sucess and connectivity using three complementary connectivity metrics
as outlined in \cite{Hofmann.2022}. First, we generated heatmaps depicting the
frequency at which different areas in the landscape were visited by simulated
dispersers. Such heatmaps serve to detect dispersal hotspots and areas of
intense use. However, they are less suitable for detecting pinchpoints and
bottlenecks that are critical in linking distinct patches. Hence, we also
computed spatially explicit betweenness scores which are useful to highlight
exactly such pinchpoints \citep{Bastille-Rousseau.2018, Bastille-Rousseau.2021}.
To compute betweenness, we overlayed the study area with a regular grid with 2.5
km x 2.5 km grid cells and determined how often simulated individuals
transitioned from one grid-cell to another. Note that in case the same
individual repeatedly realized the same cell-transition (e.g. repeatedly moved
between A-B-A-B...), we only counted a single transition to avoid emphasis on
regions where individuals moved in circles. With on the so generated
transitions, we generated a network using the centers of all grid-cells as nodes
and cell-transitions as weighted connections between the nodes. Based on this
network we computed weighted betweenness scores using the R-package
\textsf{igraph}. As a final connectivity metric and metric of dispersal success,
we calculated the number of successful dispersal events between the different
source areas as well as towards the emigration zones. We coin this type of
connectivity ``inter-patch connectivity'' as it relates to the movement between
distinct patches. Dispersal between two areas was said to be successful whenever
a trajectory leaving one area intersected with the target area. To gauge the
dispersal duration needed to move between patches (be consistent with
``patches'', ``source-areas'' etc.), we also recorded the minimum number of
steps that individuals moved before arriving at the respective patch.

@@

\subsection{Heatmaps}
Heatmaps produced from simulated dispersal trajectories reveal that the OD acts
as a major dispersal barrier during periods of high flood, but reveals viable
dispersal corridors during periods of low flood (\Cref{Heatmaps}). During
minimum flood, the area north-west of Maun appears to serve as vital dispersal
habitat. The same area is entirely avoided during maximum flood. Besides
striking differences in connectivity for the close vicinity of the delta, the
remainder of the study area shows only marginal differences in connectivity
between the two scenarios. For instance, in both scenarios the area south of the
Linyanti swamp appears as frequently visited dispersal habitat. Additional
heatmaps highlighting differences in connectivity for each source area
separately are provided in Appendix SX.

@@

>>>>

\section{Introduction}
\subsection{Climate Change and Dispersal}
Climate change is expected to profoundly impact ecosystems worldwide with
far-reaching consequences for species living therein \citep{Ozgul.2010,
Radchuk.2019, IPCC.2022}. By altering environmental conditions, climate change
affects animal behavior \citep{Fuller.2016}, resource availability
\citep{Durant.2007}, population dynamics \citep{Paniw.2021}, and the
distribution of wild living animal populations \citep{Thomas.2004,
Thuiller.2006}. One possible life-history pathway through which species may
mediate the negative consequences of environmental change is dispersal
\citep{Anderson.2012}, the movement of individuals away from their natal
location to the site of first reproduction \citep{Clobert.2012}. Through
dispersal, species can adapt to climate change by tracking favorable habitat
conditions \citep{Raia.2012} and by shifting into a different region of their
fundamental niche \citep{Kokko.2006}. Dispersal furthermore provides resilience
against changing environmental conditions by safeguarding genetic diversity
\citep{Frankham.2002, Leigh.2012, Baguette.2013}, facilitating the colonization
empty habitats \citep{Gustafson.1996, Hanski.1999a, MacArthur.2001}, and
promoting the reinforcement of weakened and small subpopulations
\citep{Brown.1977}. Through these processes, dispersal crucially contributes to
long term population viability \citep{Kokko.2006, Fahrig.2003}.

\subsection{Connectivity}
While dispersal offers a means to offset the negative demographic consequences
of climate change \citep{Kokko.2006, Hodgson.2009, Travis.2013}, it itself is a
function of climatic and environmental conditions (e.g. \citep{Elliot.2014,
Behr.2020}). The link between dispersal and the environment can either be
indirect, for example if the propensity of individuals to disperse depends on
environmental conditions, or direct, when the biophysical environment through
which dispersers move affects dispersal prospects \citep{Travis.2013}. The
latter case highlights that dispersal is also inextricably linked to the concept
landscape connectivity \citep{Baguette.2013}, which is understood as the degree
to which the landscape facilitates or impedes movements \citep{Taylor.1993}. A
sufficient degree of landscape connectivity is therefore a critical prerequisite
for successful dispersal \citep{Fahrig.2003}. Conservation strategies that aim
to facilitate dispersal by improving landscape connectivity are thus often
viewed as pinacle of conservation strategies to combat the adverse effects of
climate change \citep{Heller.2009}. Despite this, the continued degradation and
destruction habitats worldwide continues to imperil the dispersal ability of
many species and increases the prospects of inbreeding, genetic drift and local
extirpation due to stochastic events \citep{Melbourne.2008, Sawyer.2011}
alltogether limiting species' ability to adapt to changing environmental
conditions \citep{Fahrig.2003}. Succssful conservation or restoration of areas
that are of particular importance to dispersing animals thus demands a thorough
understanding of the factors that promote or limit dispersal
\citep{Baguette.2013}.

\subsection{Modeling Connectivity}
To study connectivity, various modeling techniques have emerged
\cite{Etherington.2016, Diniz.2019}. Initially, the techniques were limited to
studying strucutral aspects of connectivity by focusing on the composition and
configuration of habitat patches, while ignoring species' responses to the
landscape matrix \citep{Tischendorf.2000, Doerr.2011}. With the increasing
availability of telemetry data and methods to study species' habitat and
movement preferences \citep{Boyce.2002, Fortin.2009, Cushman.2010, Avgar.2016},
however, the focus has shifted towards a more functional understaning of
connectivity, which also takes into account how species interact with their
surroundings \citep{Tischendorf.2000, Doerr.2011}. Currently, the most prominent
\textit{functional} connectivity models are based on least-cost path analysis
(LCPA, \citealp{Adriaensen.2003}) and circuit theory (CT, \citealp{McRae.2008}),
two graph-based methods that estimate conductance of the landscape by means of a
resistance (or inversely permeability) surface \citep{Zeller.2012}. Since both
approaches make assumptions that are rarely met by dispersing individuals,
individual-based movement models (IBMMs) have also gained some momentum recently
\citep{Kanagaraj.2013, Allen.2016, Hauenstein.2019, Diniz.2019, Zeller.2020,
UnnithanKumar.2022, UnnithanKumar.2022a, Hofmann.2022}. In these models,
dispersal of individuals is explicitly simulated across the landscape of
interest and connectivity metrics are derived based on the simulated dispersal
paths \citep{Diniz.2019}. Because movement is simulated spatially and temporally
explicitly, IBMMs provide great modeling flexibility and are considered powerful
tools for examining connectivity under different landscape configurations
\citep{Littlefield.2019, UnnithanKumar.2022a}.

\subsection{Seasonality or Climate Change}
While most studies examine connectivity assuming a static landscape, this hardly
ever captures the true nature of most ecosystems, which typically experience
dynamic environmental change, either due to seasonality or climatic change.
Connectivity analyses that omit such dynamism not only fail to capture an
important aspect of reality, but also prohibit studying the sensitivity of
connectivity with regards to environmental change, such as the one expected
under predicted climate change. Recent studies that have incorporated seasonal
or climatic dynamism in their analysis have revealed vast differences in
connectivity. For instance, \cite{Mui.2017} developed seasonal resistance maps
for Blanding's turtle \textit{Emydoidea blandingii} and showed that connectivity
was substantially lower in late summer compared to spring. Similarly,
\cite{Osipova.2019} studied connectivity for African elephants
(\textit{Loxodonta africana}) during wet and dry season, and found that ignoring
seasonality resulted in an underestimation of connectivity during the wet
season, but an overestimation during the dry season. For the same species,
\citep{Kaszta.2021} generated monthly connectivity maps, demonstrating stark
differences in connectivity across the year. Other studies have used
species-distribution models in conjuction with climate predictions to highlight
how connectivity might change in result to species shifting their ranges. For
example, \cite{Ashrafzadeh.2019} found that cimate change is likely to lead to
increased habitat fragmentation and therefore reduced connectivity for mountain
newts in Iran. Similarly, \cite{Luo.2021} predicted significant reductions in
connectivity for the giant spiny frog in South-East Asia. Climate change may
thus lead to the disruption of existing dispersal corridors but may also provide
possibilities to access previously unsuitable dispersal routes. Designing and
implementing cost-effective conservation strategies that aim to improve
landscape connectivity, such as habitat restoration or the creation of wildlife
corridors, requires better understanding of such dynamics.

\subsection{What we Did}
Here, we capitalize on naturally occuring seasonal changes to approximate the
likely changes in connectivity that will arise under amplified climatic
conditions due to climate change. Specifically, flood-cycles of the OD. While a
better understanding of the ODs flooding regime under global change is of major
interest, reliable predictions remain challenging due to the complex interplay
between climate, anthropogenic water use, animal behavior, and topography of the
region.

@@

@@

The Okavango delta in Southern Africa poses a unique opportunity to study the
impacts of environmental change on species dispersal ability and connectivity in
a large scale natural experiment setup. Talk about flood cycles. According to
projections by the IPCC, Botswana and its surroundings are among the most
vulnerable to climate change. While global temperatures are expected to increase
by xx degrees until the end of the 21st century, For example,
\cite{Engelbrecht.2015a} predicts an increase between 4 and 6\degree C for
regions in southern Africa. One of the keyspone predators and an umberella
species for conservation efforts in this ecosystem is the African wild dog
(\textit{Lycaon pictus}). This species heavily relies on dispersal for long-term
persistence. The African wild dog is a lightly built canid lives in cohesive
packs comprising up to 20 individuals. Wild dogs are social breeders and the
majority of reproduction is monopolized by the dominant alpha couple who are
supported by subordinates in taking care of their pups. The species exhibits
strong inbreeding avoidance, hence juvenile individuals born into the pack
disperse in same-sex sibling coalitions after reaching sexual maturity
\citep{McNutt.1996, Behr.2020}. Notably, the African wild dog is the sole extant
representative of its genus and thus considered as high-priority species for
conservation efforts \citep{Leigh.2012}. However, fragmentation and destruction
of their historic habitat may have decreased dispersal success for this species
and thus jeopardizes connectivity \citep{Leigh.2012}. Since the impacts of
climate change are not uniformly distributed across the planet, but particularly
pronounced at high latitudes and arid ecosystems, the species may be seen as
most vulnerable to such ecosystem shifts.

Here, we capitalize on the previously parameterized dispersal model by
\cite{Hofmann.2023} and use IBMMs to simulate dispersal and assess dispersal
success and connectivity patterns for African wild dogs under two extreme
scenarios; one assuming maximum flooding of the Okavango delta and one assuming
minimum flooding of the Okavango delta. Given that water poses a barrier to
dispersing individuals, we anticipated that dispersal prospects and connectivity
during maximum flood are low. During minimum flood, in contrast, we expect to
reveal the presence of several dispersal corridors that may be used to freely
and quickly move between remaining habitat patches.

@@

\subsection{Spatial Habitat Layers}
We represented the physical landscape through which dispersers could move by a
set of spatially referenced habitat layers, each resolved at 250m x 250m. The
set of layers included water-cover, distance-to-water, tree-cover,
shrub/grassland-cover, and a human influence layer depicting anthropogenic
influences through villages, roads, and agriculture. A detailed description of
the different habitat layers is provided in \Cref{Hofmann.2021}. Importantly,
the water-cover and derived distance-to water layers were generated using MODIS
Terra MCD43A4 satellite imagery that was classified using a ``floodmapping''
algorithm developed by \citep{Wolski.2017} and available through the R-package
\textsf{floodmapr}. The algorithm allowed us to generate almost weekly updated
``floodmaps'', thus providing detailed information about the flood-extent at any
given point in time. In total, we generated 700 floodmaps between the years 2000
and 2019. Based on these maps, we generated a minimal and maximum flood
scenario. To create the minimum flood scenario, we averaged the 50 floodmaps
with smallest flood extent and generated a binary image using areas that were
inundated in at least 50\% of the maps. The resulting final maps are depicted in
\Cref{FloodExtent}.

@@

\subsection{Dispersal Model}
Our dispersal model was based on a previously parametrized and validated
integrated step-selection function (iSSF, \citealp{Avgar.2016}) fitted to GPS
data of dispersing AWDs \citep{Hofmann.2023}. In step selection functions (SSFs,
\citealp{Fortin.2005}), observed GPS locations are converted into steps (the
straight-line traveled between two GPS recordings \citep{Turchin.1998a}) and
compared to a set of \textit{random} steps in a (mixed effects) conditional
logistic regression framework \citep{Fortin.2005, Thurfjell.2014, Muff.2020,
Fieberg.2021}. The model presented in \citep{Hofmann.2023} used dispersal data
of 16 dispersing African wild dogs from a free-ranging wild dog population in
northern Botswana. GPS data during dispersal was collected at 4-hourly intervals
and translated into steps of similar duration. Observed steps were then paired
step with 24 random steps that were generated using a uniform distribution for
turning angles (\(-\pi, +\pi\)) and step lengths from a gamma distribution
fitted to observed steps (scale \(\theta\) = 6'308 and shape \(k\) = 0.37). It
was then assumed that animals assigned to each observed and random step a
selection score of the form \citep{Fortin.2005}:

@@

\subsection{Source Areas and Emigration Zones}
We simulated dispersing AWDs originating from six distinct source areas located
in the vicinity of the Okavango delta (\Cref{StudyArea}). For areas one to six
we selected locations at the delta's periphery that remained dry during minimum
and maximum flooding. At those locations, we generated circular buffers with a
radius of 20 km. For source area six, we isolated a polygon covering Chief's
Island, a peninsula located at the OD's center. Besides source areas, we also
generated ``zones of emigration'' that we used as checkpoints to determine if
and where simulated individuals left the delta's vicinity (\Cref{StudyArea}). We
generated emigration zones by first overlaying the OD with an elliptic buffer
zone that we dissected into roughly equally sized pieces in accordance to
cardinal points (\Cref{StudyArea}).

\subsection{Dispersal Simulation}
Originating from each source area we simulated 1'000 individuals dispersing for
a total of 2000 steps (which equals to a dispersal duration of 400 days). Each
individual was simulated once assuming a minimum flood, once assuming a maximum
flood. This resulted in the simulation of a total of 12'000 individuals. The
simulation algorithm was based on the algorithm described in \Cref{Hofmann.2023}
and works as follows. A random location within the source area is chosen as a
starting point. Originating from the starting point, a set of 25 random steps is
generated by sampling step lengths from a gamma distribution fitted to observed
steps (what is a step?) (shape = , scale = ) and turning angles from a uniform
distibution ($-\pi, +\pi$). Along each random step the underlying spatial
covariates are extracted and relevant movement metrics are computed (i.e.
log(sl), cos(ta), ta). The parametrized dispersal model is used to predict the
probability of each step for being chosen given the steps covariate values. One
of the steps is sampled based on assigned probabilities and the location of the
animal is updated. The procedure is then repated until the desired number of
steps is realized. Here, we simulated each individual for 2'000 steps, which
correspons to the longest dispersal duration recorded in our data. The original
model was trained using 4-hourly steps, thus a simulated step also resembled the
movement conducted within four hours. Trajectories resulting from such a
simulation can be understood as correlated random walks that take into account
both habitat and movement preferences of dispersing individuals.

@@

\subsection{Dispersal Prospects and Connectivity and Distance To Humans}
Based on simulated dispersal trajectories we quantified dispersal sucess and
connectivity using three complementary connectivity metrics as outlined in
\cite{Hofmann.2023}. First, we generated heatmaps depicting the frequency at
which different areas in the landscape were visited by simulated dispersers.
Such heatmaps serve to detect dispersal hotspots and areas of intense use.
However, they are less suitable for detecting pinchpoints and bottlenecks that
are critical in linking distinct patches. Hence, we also computed spatially
explicit betweenness scores which are useful to highlight exactly such
pinchpoints \citep{Bastille-Rousseau.2018, Bastille-Rousseau.2021}. To compute
betweenness, we overlayed the study area with a regular grid comprising of 2.5
km x 2.5 km grid cells and determined how often simulated individuals
transitioned from one grid-cell to another. To determine such transitions, we
assessed the sequence of grid-cells thorugh which each individual moved. In case
the same individual repeatedly realized the same cell-transition (e.g.
repeatedly moved between A-B-A-B...), we only counted a single transition to
avoid emphasis on regions where individuals moved in circles. With on the so
generated transitions, we generated a network using the centers of all
grid-cells as nodes and cell-transitions as weighted connections between the
nodes (for details, see \citep{Hofmann.2023}). Based on this network we computed
weighted betweenness scores. As a final connectivity metric, we quantified
dispersal success and calculated the number of successful dispersal events
between the different source areas. We coin this type of connectivity
``inter-patch connectivity'' as it relates to the movement between distinct
patches. Dispersal between two areas was said to be successful whenever a
trajectory leaving one area intersected with the target area. We also tallied
the number of individuals that left the OD's vicinity and moved through one of
the emigration zones. To gauge the dispersal duration needed to move between
patches (be consistent with ``patches'', ``source-areas'' etc.), we recorded the
minimum number of steps that individuals moved before arriving at the respective
patch. Finally, we identified zones with a high potential for human wildlife
conflict. For this, we isolated all simulated locations that lied within a xx km
buffer to the nearest human-influenced pixel in the study area. Based on the so
isolated coordinates we generated heatmaps highlighting areas with a high
density of coordinates within the vicinity of humans.

@@

\subsection{Heatmaps}
Heatmaps produced from simulated dispersal trajectories reveal that the OD acts
as a major dispersal barrier during periods of high flood, but reveals viable
dispersal corridors during periods of low flood (\Cref{Heatmaps}). During
minimum flood, the area north-west of Maun appears to serve as vital dispersal
habitat, yet the same area is entirely avoided during maximum flood. Besides
striking differences in connectivity for the close vicinity of the delta, the
remainder of the study area shows only marginal differences in connectivity
between the two scenarios. For instance, in both scenarios the area south of the
Linyanti swamp appears as frequently visited dispersal habitat. Additional
heatmaps highlighting differences in connectivity for each source area
separately are provided in Appendix SX.

@@

The density around Maun increases from xx to xx, the density around the pan
handel from to xx
